raptr: Representative and Adequate Prioritisation Toolkit in R

### 2016-11-11 code/rmarkdown/article.Rmd
+ added feedback after talk with Hugh

1. An underlying aim in conservation planning is to maximise the long-term persistence of biodiversity. To fulfill this aim, the processes that sustain biodiversity must be conserved. Many biodiversity processes can be preserved by conserving a representative sample of each biodiversity feature (species or ecosystem). For example, representing ability for individuals to persist in a range of climatic conditions might be crucial for the long-term persistence of a species in a world where environmental change is accelerating. Yet current approaches overwhelmingly focus on securing a target proportion of features' geographic ranges; little attention is directed towards targeting a representative sample of them.

2. To address this issue, we developed the \\texttt{raptr: Representative and Adequate Prioritisation Toolkit in R} to guide reserve selection using explicit targets for representing biodiversity processes as well as area protection targets. Users set \"space targets\" to secure a representative sample of each feature within an attribute space that reflects a biodiversity process (eg. an attribute space expressing variation in genetic characteristics between individuals, or variation in climatic conditions between areas). Users also set \"amount targets\"--similar to conventional reserve selection methods--to secure a target proportion of the feature's geographic distribution. We demonstrated the functionality of this package using a simulated study and two case studies. In each study, we generated solutions that secured a proportion of the species' distributions using amount targets--representing conventional reserve selection--and compared them to solutions generated using both amount and space targets.

3. We show that explicitly considering biodiversity processes in reserve selection can substantially change the solution. The simulations suggest that including representativeness in conservation planning is important where biodiversity features have multimodal distributions in an attribute space. The case studies show that explicitly considering space-based targets for biodiversity processes can result in solutions that could be far more effective in achieving long term biodiversity persistence and do not require a much greater area reserved than traditional approaches.

@@

<!-- biodiversity processes are important -->
Prioritisations must capture the ecological and evolutionary processes that sustain biodiversity [@r3; @r255; @r9]. Ecological processes are important for biodiversity to persist over short time-scales. For instance, predator-prey processes can mediate ecosystem structure and function through top-down control [@r476]. Additionally, migratory processes allow many species to access suitable habitat throughout the year [@r477]. On the other hand, evolutionary processes are important for biodiversity to persist over long time-scales. Adaptive evolutionary processes foster resilience against environmental change [@r3; @r43], and neutral evolutionary processes maintain genetic diversity [@r5; @r67; @r63]. In recent decades, a wealth of data relevant to biodiversity processes has become available [eg. bioclimatic, distribution, genetic, phylogenetic, and trait data; @r33; @r460; @r478; @r453]. This has sparked significant interest in using this data to make better conservation decisions [@r68; @r9; @r473; @r247; @r448; @r468; @r2; @r407; @r61; @r470; @r475; @r469].

<!-- current methods for preserving biodiversity processes -->
Today, the most widely used conservation planning tools focus on biodiversity patterns [\texttt{Marxan} and \texttt{Zonation}; @r429; @r22]. The cornerstone of these tools is that they allow decision makers to set targets to obtain solutions that secure a proportion of the geographic range of biological features (populations, species, or ecosystems of conservation interest). To incorporate some information on biodiversity processes, current methods involve partitioning the features into different groups as a pre-processing step [@r9; @r247; @r448]. These groups are partitioned based on a proxy variable that relates to a biodiversity process of interest [@r473]. For instance, species ranges' have been partitioned according to habitat discontinuities to capture adaptive processes [@r2]. However, by partitioning features into different groups, it is not clear how well a given solution samples the original proxy variable, and ultimately, how well the solution captures the biodiversity process of interest. This limitation is also present in some of the earliest reserve selection methods [eg. @r295], and to overcome it, Faith and Walker [-@r218] developed a decision support tool (\texttt{DIVERSITY}) to explicitly sample proxy variables to secure biodiversity processes. However, this tool is not widely used today, in part, because it cannot accommodate explicit targets on biodiversity patterns for multiple features (unlike and \texttt{Marxan} and \texttt{Zonation}).

<!-- aims of this paper -->
One of the key issues in reserve selection is the lack of a unified decision support tool that uses explicit and separate targets for biodiversity patterns and processes. To begin to fill this gap, we united the ideas underpinning \texttt{DIVERSITY} and \texttt{Marxan} into new formulations of the reserve selection problem, and implemented them in the \texttt{raptr R} package. This \texttt{R} package provides decision makers with the tools to generate prioritisations that preserve both biodiversity patterns and processes. Here, we aim to the provide an in-depth understanding of this \texttt{R} package and explore its functionality.

# Methods
## OPTIMISATION PROBLEM FORMULATION
Biodiversity features are defined as the entities that the prioritisation is required to preserve (eg. species, ecosystems). Spatial attributes are defined as the variation across the feature' range that the prioritisation is required to sample. They can be intrinsic (eg. genetic or trait variation) or extrinsic (eg. environmental variation) to the feature. These attributes should be related to the biodiversity processes that the decision maker aims to preserve.

@@

Demand points are designated by the decision maker to indicate regions of the attribute space that they wish to represent in the prioritisation (see below for discussion on how demand points can be generated for real-world data sets). The amount of variation in the attribute space that a solution secures depends on the distance between each demand point and each selected planning unit in the attribute space (see below for mathematical expressions). The shorter the distances between the demand points and the planning units; the better the prioritisation is at securing the variation in the spatial attribute. To calculate the proportion of variation secured in a given solution, the distances are scaled by the distance between the demand points to their centroid (equivalent to calculating R$^2$ values for K-means clustering). In any attribute space there may exist points that are impossible (eg. mean annual rainfall -5 mm), or do not occur in the study area (eg. mean annual temperature 30$^{\circ}$C in Antarctica). Additionally, there may be some regions that are desirable for some features and undesirable for others (eg. conditions known to be outside the physiological tolerance of certain species). Thus a different set of demand points and weights are used for each attribute space and each feature. By using demand points in desirable regions of an attribute space for a given feature, the decision maker can ensure that prioritisations secure the feature in planning units with spatial attributes that are desirable for that feature.

To illustrate these concepts, consider the following example: we wish to develop a prioritisation for a single species that has four populations. However, we can only afford to preserve three of the four populations. To ensure that we preserve a representative sample, we want to preserve populations inhabiting different environmental conditions. We obtained annual rainfall (mm) and temperature ($^{\circ}$C) data at the location of each population. We used this data to construct a two-dimensional environmental attribute space. Next, we generated demand points as equi-distant points inside this space. By comparing the distributions of demand points and populations in the attribute space, we can identify a suitable prioritisation (Fig. 1). We can see that a solution that prioritises both populations _A_ and _C_, effectively "doubles-up" on the same environmental characteristics and would contain in considerable redundancy. Instead, a more representative sample of the intra-specific variation could be captured by securing populations $A$, $B$, and $D$. This example demonstrates how attribute spaces and demand points relate to biodiversity processes, and can be used to guide the reserve selection process.

The formulations used to express the reserve selection problem in the \texttt{raptr R} package are based on a combination of the \texttt{Marxan}, \texttt{DIVERSITY}, and the uncapacitated facility location problems [@r451]. Although the \texttt{raptr R} package utlises two novel formulations, for brevity, we will define the simpler formulation--referred to as the unreliable formulation--below and define the more complex version--the reliable formulation--in Appendix S1. These formulations are based on unreliable and reliable facility location problems and have been named accordingly [@r16]. The key difference between these two formulations is that the reliable formulation explicitly considers the probability that planning units are occupied by features when calculating the proportion of an attribute space sampled in a solution. All mathematical terms defined hereafter are described in Table S1. For convenience, the cardinality of sets will be denoted using the same symbol used to denote the set.

@@

To understand the behaviour of the unreliable problem and showcase its value, we conducted a simulation study and two case studies. These studies involved generating solutions using only amount targets to represent prioritisations generated using conventional methods (eg. \texttt{Marxan}), and solutions using amount and space targets using the unreliable formulation. By comparing these solutions, we can guage the benefits of explicitly including space targets in reserve selection. All analyses were performed in \texttt{R} [version `r getRversion()`; @r192].

@@

### Results
Generally, the solution generated using amount targets preserved a representative sample of each the `r as.character(as.english(length(cs1.params.LST[[MODE]]$species.names)))` bird species' niches--with `r ifelse(length(amount.not.represented.species.names)>1, as.character(as.english(length(amount.represented.species.names))), 'one')` excpetion. Specifically, this solution sampled over `r cs1.params.LST[[MODE]]$space.target*100` % of the realised niche of `r parsed.representative.space.held.names`. Yet it failed to achieve this for `r parsed.not.representative.space.held.names`. This result demonstrates that though solution generated using conventional methods may preserve a representative sample of some features (Figure S1), only through the use of explicit targets can decision makers be sure that solutions secure a representative sample for all features.

@@

To maximise the long-term persistence of biodiversity, decision makers need to identify prioritisations that preserve existing patterns of biodiversity and the processes that support them. To achieve this, conservation planners need a decision support tool that can explicitly accommodate biodiversity patterns and processes. Here, we developed the \texttt{raptr R} package to fill this void. By exploring the functionality of this package using several simulated and case study species, we found that including space-based targets can result in substantially better solutions.

# Acknowledgements
JOH is funded by an Australian Postgraduate Award (APA) scholarship. RAF has an Australian Research Council Future Fellowship. This work was supported by the Centre of Biodiversity and Conservation Science (CBCS).

>>>>

1. An underlying aim in conservation planning is to maximise the long-term persistence of biodiversity. To fulfill this aim, the ecological and evolutionary processes that sustain biodiversity must be preserved. One way to conserve such processes at the feature level (eg. species, ecosystem) is to preserve a representative sample of the physical attributes that underpin them across the feature's geographic distribution. For example, preserving the adaptive processes currently acting on a species might be crucial for its long-term persistence in a world where environmental change is accelerating. By conserving individuals with the ability to persist in a range of climatic conditions--physical attributes associated with adaptation--protected areas can foster evolutionary processes. Yet current approaches overwhelmingly focus on securing a target proportion of each features' geographic range; little attention is directed towards targeting a representative sample of them.

2. To address this issue, we developed the \\texttt{raptr: Representative and Adequate Prioritisation Toolkit in R} to guide reserve selection using explicit targets for representing biodiversity processes as well as more typical area protection targets. Users set \"amount targets\"--similar to conventional reserve selection methods--to secure a target proportion of the feature's geographic distribution. Additionally, users set \"space targets\" to ensure that solutions also secure a representative sample of an evolutionary or ecologically relevant attribute (space) across a feature's geographic distribution (eg. an attribute expressing variation in genetic characteristics between individuals, or variation in climatic conditions between areas). We demonstrate the functionality of this package using simulations and two case studies. We generate solutions that secure a proportion of the species' distributions using amount targets--representing conventional reserve selection--and compare them with solutions generated using both amount and space targets.

3. We show that markedly different solutions emerge when explicitly considering within feature representation in reserve selection. Our simulations suggest that including representativeness in conservation planning is important where biodiversity features have multimodal distributions in an attribute space. The case studies show that explicitly setting space-based targets can result solutions that preserve a much more representative sample of features. Additionally, the case studies suggest that securing a representative sample of features may not require a much  greater area of reservation than traditional approaches.

@@

<!-- biodiversity processes are important -->
Prioritisations must capture the ecological and evolutionary processes that sustain biodiversity [@r3; @r255; @r9]. Ecological processes are important for biodiversity to persist over short time-scales. For instance, predator-prey processes can mediate ecosystem structure and function through top-down control [@r476]. Additionally, migratory processes allow many species to access suitable habitat throughout the year [@r477]. On the other hand, evolutionary processes are important for biodiversity to persist over long time-scales. Adaptive evolutionary processes foster resilience against environmental change [@r3; @r43], and neutral evolutionary processes maintain genetic diversity [@r5; @r67; @r63]. In recent decades, a wealth of data relevant to biodiversity processes has become available [eg. bioclimatic, distribution, genetic, phylogenetic, and trait data; @r33; @r460; @r478; @r453]. This has sparked significant interest in using these data to make better conservation decisions [@r68; @r9; @r473; @r247; @r448; @r468; @r2; @r407; @r61; @r470; @r479; @r475; @r469].

<!-- current methods for preserving biodiversity processes -->
Today, the most widely used conservation planning tools focus on biodiversity patterns [\texttt{Zonation} and \texttt{Marxan}; @r429; @r22]. These tools allow decision makers to obtain solutions that secure a proportion of the geographic range of biological features (populations, species, or ecosystems of conservation interest). To incorporate some information on biodiversity processes, current methods involve partitioning the features into different groups as a pre-processing step. The features are partitioned based on an attribute variable that relates to the biodiversity process of interest [@r9; @r247; @r448]. For instance, species distributions can be divided according to habitat discontinuities to capture adaptive processes [@r2]. However, using this method is challenging because biodiversity data are often continuous and multi-dimensional.

<!-- issues with partitioning features -->
After partitioning features based on continuous variables and obtaining a solution, it is not clear how well the solution samples the original continuous variables. For instance, if a feature was split into four sub-features based on a single continuous variable, and a solution samples 10 % of each sub-feature, how much of the original continuous variation does the solution sample? This is not immediately clear. If the solution sites reserves evenly along the original variable then it may sample far more than 10 % of the original variation. Whereas, if the solution sites reserves near the values used to divide the continuous values [as observed when partitioning continuous geographic variation based on political boundaries; @r483], then it may sample much less of the original variation. This limitation has not arisen recently. In fact, it dates back to some of the earliest reserve selection methods [eg. @r295].

<!-- diversity method for conservation -->
During the early 1990s, Faith and Walker [-@r480] developed a decision support tool (\texttt{DIVERSITY}) to identify prioritisations that secure a representative sample of continuous environmental variation across a study area. Although this method has been pitched as an alternative to species-based conservation planning [@r481; @r218; @r203], it can also identify solutions that conserve a representative sample of variation across a single species' range. Recent work has built on this reserve selection method by solving problems with more advanced optimisation routines [@r482]. However, \texttt{DIVERSITY}-based reserve selection methods are not widely used today, in part, because they do not accommodate explicit amount-based targets for multiple features (_cf_. \texttt{Marxan} and \texttt{Zonation}).

<!-- aims of this paper -->
One of the key issues in reserve selection is the lack of a unified decision support tool that uses explicit and separate targets to obtain solutions that secure an adequate amount of habitat for each feature and a representative sample of each feature. To begin to fill this gap, here we unite the ideas underpinning \texttt{DIVERSITY} and \texttt{Marxan} into new formulations of the reserve selection problem, and implement them in the \texttt{raptr R} package. This \texttt{R} package provides decision makers with the tools to generate prioritisations based on data that relates to biodiversity patterns and processes. Here, we aim to the provide an in-depth understanding of this \texttt{R} package and explore its functionality.

# Methods
## OPTIMISATION PROBLEM FORMULATION
Biodiversity features are defined as the entities that the prioritisation is required to preserve (eg. species, ecosystems). Spatial attributes are defined as the variation across the feature' range that the prioritisation is required to sample. They can be intrinsic variation (eg. genetic or trait) variation that is extrinsic to the feature (eg. environmental conditions). There should, however, be a reasonable underlying hypothesis that relates this variation to the biodiversity processes that the decision maker aims to preserve.

@@

Demand points [@r480; @r218; @r203] are designated by the decision maker to indicate regions of the attribute space that they wish to represent in the prioritisation (see below for discussion on how demand points can be generated for real-world data sets). The amount of variation in the attribute space that a solution secures depends on the distance between each demand point and each selected planning unit in the attribute space (see below for mathematical expressions). The shorter the distances between the demand points and the planning units; the better the prioritisation is at securing the variation in the spatial attribute. To calculate the proportion of variation secured in a given solution, the distances are scaled by the distance between the demand points to their centroid (equivalent to calculating R$^2$ values for K-means clustering). In any attribute space there may exist points that are impossible (eg. mean annual rainfall -5 mm), or do not occur in the study area (eg. mean annual temperature 30$^{\circ}$C in Antarctica). Additionally, there may be some regions that are desirable for some features and undesirable for others (eg. conditions known to be outside the physiological tolerance of certain species). Thus a different set of demand points and weights are used for each attribute space and each feature. By using demand points in desirable regions of an attribute space for a given feature, the decision maker can ensure that prioritisations secure the feature in planning units with spatial attributes that are desirable for that feature.

To illustrate these concepts, consider the following example: we wish to develop a prioritisation for a single species that has four populations. However, we can only afford to preserve three of the four populations. To ensure that we preserve a representative sample, we want to preserve populations inhabiting different environmental conditions. We obtained annual rainfall (mm) and temperature ($^{\circ}$C) data at the location of each population. We used this data to construct a two-dimensional environmental attribute space. Next, we generated demand points as equi-distant points inside this space. By comparing the distributions of demand points and populations in the attribute space, we can identify a suitable prioritisation (Fig. 1). We can see that a solution that prioritises both populations _A_ and _C_, effectively "doubles-up" on the same environmental characteristics and constitute considerable redundancy if both selected. Instead, a more representative sample of the intra-specific variation could be captured by securing populations $A$ (or $C$), $B$, and $D$. This example demonstrates how attribute spaces and demand points relate to biodiversity processes, and can be used to guide the reserve selection process.

The formulations used to express the reserve selection problem in the \texttt{raptr R} package are based on elements of the \texttt{Marxan}, \texttt{DIVERSITY}, and the uncapacitated facility location problems [@r451]. Although the \texttt{raptr R} package utlises two novel formulations, for brevity, we will define the simpler formulation--referred to as the unreliable formulation--below and define the more complex version--the reliable formulation--in Appendix S1. These formulations are based on unreliable and reliable facility location problems and have been named accordingly [@r16]. The key difference between these two formulations is that the reliable formulation explicitly considers the probability that planning units are occupied by features when calculating the proportion of an attribute space sampled in a solution. All mathematical terms defined hereafter are described in Table S1. For convenience, the cardinality of sets will be denoted using the same symbol used to denote the set.

@@

To understand the behaviour of the unreliable problem and showcase its value, we conducted a simulation study and two case studies. These studies involved generating solutions using only amount targets to represent prioritisations generated using conventional methods (eg. \texttt{Marxan}), and solutions using amount and space targets using the unreliable formulation. By comparing these solutions, we can gauge the benefits of explicitly including space targets in reserve selection. All analyses were performed in \texttt{R} [version `r getRversion()`; @r192].

@@

### Results
Generally, the solution generated using amount targets preserved a representative sample of each the `r as.character(as.english(length(cs1.params.LST[[MODE]]$species.names)))` bird species' niches--with `r ifelse(length(amount.not.represented.species.names)>1, as.character(as.english(length(amount.represented.species.names))), 'one')` excpetion. Specifically, this solution sampled over `r cs1.params.LST[[MODE]]$space.target*100` % of the realised niche of `r parsed.representative.space.held.names`. Yet it failed to achieve this for `r parsed.not.representative.space.held.names`. This result demonstrates that though solutions generated using conventional methods may preserve a representative sample of some features (Figure S1), only through the use of explicit targets can decision makers be sure that solutions secure a representative sample for all features.

@@

To maximise the long-term persistence of biodiversity, decision makers need to identify prioritisations that preserve existing patterns of biodiversity and the processes that support them. To achieve this, conservation planners need a decision support tool that can explicitly accommodate data on biodiversity patterns and processes. Here, we developed the \texttt{raptr R} package to fill this void. By exploring the functionality of this package using several simulated and case study species, we found that including space-based targets can result in substantially different solutions.

# Acknowledgements
JOH is funded by an Australian Postgraduate Award (APA) scholarship. RAF has an Australian Research Council Future Fellowship. This work was supported by the Centre for Biodiversity and Conservation Science (CBCS).
